---
title: "Bar Plots with Multiple Groups and Error Bars"
format: gfm
prefer-html: true
self-contained: false
editor: visual
knitr:
  opts_chunk:
    comment: ""
df-print: kable
---

This is a step-by-step guide to building a bar plot using `ggplot2`. While you would typically write all components in a single code block using `+` to connect elements, we hope splitting the code will illustrate how each snippet contributes to the final visualization.

## Load libraries

```{r setup}
#| echo: true
#| output: false
library(ggplot2)
library(apde.graphs)
library(data.table)
library(scales) # to label axes with $, %, etc. 
```

## Import & preview synthetic data

```{r generate_data}
dt <- apde.graphs::icecreamDT
head(dt)
```

## Create the base `ggplot2` bar plot

```{r base_plot}
ice_cream_plot <- ggplot(dt, aes(x = factor(year), 
                                 y = mean_income,
                                 fill = flavor)) +  # different color for each flavor
  # Basic bar plot
  geom_col(position = position_dodge(width = 0.8),  # adjust horizontally to show groups
           width = 0.7)                             # adjust bar width
```

```{r display_base_plot}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add error bars to show uncertainty (95% confidence intervals)

```{r error_bars}
ice_cream_plot <- ice_cream_plot + 
  geom_errorbar(aes(ymin = mean_income - 1.96*se,       # lower bound CI
                    ymax = mean_income + 1.96*se),      # upper bound CI
                position = position_dodge(width = 0.8), # adjust horizontally to align with bars
                linewidth = 1,                          # width of the error bar
                width = 0)                              # width of the horizontal lines at the ends of the bar
```

```{r display_error_bars}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add colors and scales

```{r add_scales}
ice_cream_plot <- ice_cream_plot +
  # Custom colors using colorbrewer
  scale_fill_brewer(palette = "Dark2") +  # One of many colorblind-friendly palettes

  # Customize y-axis with dollar signs
  scale_y_continuous(
    limits = c(0, 100000),                # fix y-axis range
    breaks = seq(0, 100000, 10000),       # show tick marks every $10,000
    labels = scales::label_currency(prefix = '$')
  )
```

```{r display_colors_scales}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add labels

```{r add_labels}
ice_cream_plot <- ice_cream_plot +
  labs(
    title = 'Mean Income by Ice Cream Preference',
    subtitle = 'Non-representative Sample of Megalopolis Residents',
    x = 'Year',
    y = 'Mean Income (2020 dollars)', 
    fill = 'Favorite Flavor'  # legend title
  )
```

```{r display_labels}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add APDE customizations

The `apde_caption()` and `apde_theme()` elements are from the `apde.graphs` package, not `ggplot2`.

```{r add_theme}
#| echo: true
#| output: false
ice_cream_plot <- ice_cream_plot +
  
  apde_caption(data_source = 'Synthetic dataset') +
  
  apde_theme()
```

```{r display_apde}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add horizontal reference line

```{r add_hline}
ice_cream_plot <- ice_cream_plot +
  # Add reference line at overall mean
  geom_hline(yintercept = mean(dt$mean_income), # geom_vline for vertical line
             linetype = 'dashed',               # options: solid' dashed, dotdash, longdash, twodash
             linewidth = 1,                     # thickness of line
             color = 'red')                     # color of line
```

```{r display_hline}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Add annotations

```{r add_annotations}
ice_cream_plot <- ice_cream_plot +
  # Add explanatory text
  annotate(
    geom = 'text',
    x = 5,
    y = 95000,
    label = "Steepest increase\nfor 'Other' flavor",
    hjust = 1  # horizontal justification
  ) + 
  
  # Add arrow pointing to steepest increase
  annotate(
    geom = "segment",
    x = 4.9, xend = 5.1,
    y = 90000, yend = 85000,
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  
  # Add a shaded box around the text
  annotate(
    geom = 'rect',        # highlight an area of your graph
    xmin = 4, xmax = 5.2, 
    ymin = 90000, ymax = 100000,
    alpha = 0.2,          # transparency (0 == most, 1 == least)
    fill = 'green',       # color of box 
    col = NA)             # color of box outline
```

```{r display_annotated}
#| fig-width: 10 
#| fig-height: 6 
#| dpi: 150 
#| message: false
#| echo: false

ice_cream_plot
```

## Save the plot

`ggsave` automatically detects file type from filename extension (.jpg, .png, .pdf, etc.)

```{r save_plot}
#| echo: true
#| output: false
#| eval: false
ggsave(filename = 'income_by_ice_cream_preference.jpg', # filename with extension
       plot = ice_cream_plot,                           # plot to save
       width = 10,                                      # width in inches
       height = 6,                                      # height in inches
       dpi = 600)                                       # resolution in dots per inch
```

-- *`r paste0("Updated by ", Sys.getenv("USERNAME"), ", ", Sys.Date())`*
